Regularised parameter selection in the non-stationary Hawkes process

### 2017-05-01 _paper_output/chapter_sparse_hawkes.tex
macros

@@

@@

We are given a process
\(N:\mathcal{B}(\mathbb{R})\mapsto \mathbb{N}^0\), which counts the
number of arrivals in a given time. Since we are usually concerned with
arrivals over increasing time intervals, we will usually consider \(N\)
to be indexed by time: \(N(t):=N([0,t))-N((-\infty, t))\).

We represent a realisation of the time series by a (finite by
assumption) sequence of arrival times
\(\mathbf{t}=\{\dots,t_1, t_2,\dots,t_N(T)\}.\) We will be concerned
with estimating its likelihood based on the likelihood of the series on
an interval an interval \([0,T]\). We define
\(\mathbf{t}':=\mathbf{t}\cap [0,T]\) and assume
\(\left|\mathbf{t}'\right|<\infty.\)

We set \(t_1\equiv \inf \mathbf{t}'\); Indices \(i\leq 0\) are reserved
for point in the "history" of the process before time \(0\), which may
be generated by any arbitrary process. We will not be concerned with
evaluating the likelihood of such histories but may retain them to
control for their influence on the future of the series. For the rates
to be well-defined, however, we will assume that the increments of the
process are a.s. finite on any finite interval.

    We use only the left-continuous filtration
\(\mathcal{F}(\{N(s)\}_{s < t})\) generated by the process, which we
here write as \(\mathcal{F}(N(t)).\)

We call it a regular point process if it has an associated
\emph{intensity} process \(\lambda\), such that \(\lambda\) is
predictable with respect to \(\mathcal{F}(N)\) generated by the process,
i.e. \cite{daley_introduction_2003}

\[\lambda(t|\mathcal{F}(N(t))):=\lim_{h\to 0} \frac{\mathrm E\left(N([t,t+h))\right)}{h}.\]

Hereafter we define \(\lambda^*(t):=\lambda(t|\mathcal{F}(N(t)))\) for
brevity.

@@

From the definition of the intensity we find \[\begin{aligned}
\frac{d}{dt}P(t_{i+1}=t|\mathcal{F}(N(t)))=
\end{aligned}\]

@@

\[
\begin{aligned}
\lambda(t|\mathcal{F}(N(t);\mu,\phi) &= \mu(t) + \phi * N\\
&= \mu(t)  + \int_{-\infty}^{\infty}\phi(t-s)dN(s)\\
&= \mu(t) + \int_{-\infty}^{t}\phi(t-s)dN(s)\\
&= \mu(t) + \sum_{t_i\lt t}\phi(t-t_i)
\end{aligned}
\]

@@

\[\begin{aligned}L(N;\mu,\phi) &=-\int_0^T\lambda^*(t)dt + \int_0^T\log \lambda^*(t) dN_t \\
&=-\int_0^T\lambda^*(t)dt + \int_0^T\log \lambda^*(t) dN_t \\
&=-\int_0^T\lambda^*(t)dt + \sum_{t\in\mathbf{t}\cap [0,T]}\log \lambda^*(t)
\end{aligned}\]

Using the method of maximum likelihood we estimate our parameter vector
\(\theta:=(\mu, \phi)\) by choosing the estimates to maximise the
likelihood,
\[\hat{\theta}_\pi(N) = \operatorname{argmax}_{\mu,\phi} L_\pi(N;\theta).\]

    Here we extend the likelihood approach, treating it as penalised
functional regression problem. We augment the log-likeihood function
with a penalty functional \(\pi\):

@@

\[\hat{\theta}_\pi(N) = \operatorname{argmax}_\theta L_\pi(N;\theta)\]

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_13_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

\[\begin{aligned}
{ \scriptstyle \frac{\partial}{\partial\gamma} } L_\pi(N;\theta)&=-\int_0^T{ \scriptstyle \frac{\partial}{\partial\gamma} }\lambda^*(t;\theta)dt + \int_0^T { \scriptstyle \frac{\partial}{\partial\gamma} }\log \lambda^*(t;\theta) dN_t -{ \scriptstyle \frac{\partial}{\partial\gamma} }\pi(\theta) \\
&=-\int_0^T{ \scriptstyle \frac{\partial}{\partial\gamma} }\lambda^*(t;\theta)dt + \sum_{0\leq t_i\leq T} \frac{{ \scriptstyle \frac{\partial}{\partial\gamma} }\lambda^*(t_i)}{\lambda^*(t_i)} -{ \scriptstyle \frac{\partial}{\partial\gamma} }\pi(\theta)
\end{aligned}
\]

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_29_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_29_1.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_29_2.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_31_0.png}
    \end{center}
    { \hspace*{\fill} \\}

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_33_0.png}
    \end{center}
    { \hspace*{\fill} \\}

>>>>

\newcommand{\var}{\operatorname{Var}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\bb}[1]{\mathbb{#1}}
\renewcommand{\bf}[1]{\mathbf{#1}}
\renewcommand{\v}[1]{\boldsymbol{#1}}
\newcommand{\m}[1]{\mathrm{#1}}
\newcommand{\cc}[1]{\mathcal{#1}}
\renewcommand{\cc}[1]{\mathcal{#1}}
\newcommand{\oo}[1]{\operatorname{#1}}

@@

    \[
\renewcommand{\var}{\operatorname{Var}}
\renewcommand{\dd}{\mathrm{d}}
\renewcommand{\bb}[1]{\mathbb{#1}}
\renewcommand{\bf}[1]{\mathbf{#1}}
\renewcommand{\v}[1]{\boldsymbol{#1}}
\renewcommand{\m}[1]{\mathrm{#1}}
\renewcommand{\cc}[1]{\mathcal{#1}}
\renewcommand{\cc}[1]{\mathcal{#1}}
\renewcommand{\oo}[1]{\operatorname{#1}}
\]

@@

We are given a process \(N:\cc{B}(\bb{R})\mapsto \bb{N}^0\), which
counts the number of arrivals in a given time. Since we are usually
concerned with arrivals over increasing time intervals, we will usually
consider \(N\) to be indexed by time:
\(N(t):=N([0,t))-N((-\infty, t))\).

We represent a realisation of the time series by a (finite by
assumption) sequence of arrival times
\(\v{t}=\{\dots,t_1, t_2,\dots,t_N(T)\}.\) We will be concerned with
estimating its likelihood based on the likelihood of the series on an
interval an interval \([0,T]\). We define \(\v{t}':=\v{t}\cap [0,T]\)
and assume \(\left|\v{t}'\right|<\infty.\)

We set \(t_1\equiv \inf \v{t}'\); Indices \(i\leq 0\) are reserved for
point in the "history" of the process before time \(0\), which may be
generated by any arbitrary process. We will not be concerned with
evaluating the likelihood of such histories but may retain them to
control for their influence on the future of the series. For the rates
to be well-defined, however, we will assume that the increments of the
process are a.s. finite on any finite interval.

    We use only the left-continuous filtration
\(\mathcal{F}(\{N(s)\}_{s < t})\) generated by the paths of the process,
which we here write as \(\cc{F}(N(t)).\)

We call it a regular point process if it has an associated
\emph{intensity} process \(\lambda\), such that \(\lambda\) is
predictable with respect to \(\cc{F}(N)\) generated by the process, i.e.
\cite{daley_introduction_2003}

\[\lambda(t|\cc{F}(N(t))):=\lim_{h\to 0} \frac{\bb E\left(N([t,t+h))\right)}{h}.\]

Hereafter we define \(\lambda^*(t):=\lambda(t|\cc{F}(N(t)))\) for
brevity.

@@

From the definition of the intensity we find \[\begin{aligned}
\frac{\dd}{\dd t}\bb P(t_{i+1}=t|\mathcal{F}(N(t)))=
\end{aligned}\]

@@

\[
\begin{aligned}
\lambda(t|\mathcal{F}(N(t);\mu,\phi) &= \mu(t) + \phi * N\\
&= \mu(t)  + \int_{-\infty}^{\infty}\phi(t-s)\dd N(s)\\
&= \mu(t) + \int_{-\infty}^{t}\phi(t-s)\dd N(s)\\
&= \mu(t) + \sum_{t_i\lt t}\phi(t-t_i)
\end{aligned}
\]

@@

\[\begin{aligned}L(N;\mu,\phi) &=-\int_0^T\lambda^*(t) \dd t + \int_0^T\log \lambda^*(t) \dd N_t \\
&=-\int_0^T\lambda^*(t)\dd t + \int_0^T\log \lambda^*(t) \dd N_t \\
&=-\int_0^T\lambda^*(t)\dd t + \sum_{t\in\v{t}'}\log \lambda^*(t)
\end{aligned}\]

Using the method of maximum likelihood we estimate our parameter vector
\(\theta:=(\mu, \phi)\) by choosing the estimates to maximise the
log-likelihood,
\[\hat{\theta}_\pi(N) = \o{argmax}_{\mu,\phi} L_\pi(N;\theta).\]

    Here we extend the likelihood approach with a penalty functional
\(\pi\):

@@

\[\hat{\theta}_\pi(N) = \oo{argmax}_\theta L_\pi(N;\theta)\]

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_14_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

\[\begin{aligned}
{ \scriptstyle \frac{\partial}{\partial\gamma} } L_\pi(N;\theta)&=-\int_0^T{ \scriptstyle \frac{\partial}{\partial\gamma} }\lambda^*(t;\theta)dt + \int_0^T { \scriptstyle \frac{\partial}{\partial\gamma} }\log \lambda^*(t;\theta) \dd N_t -{ \scriptstyle \frac{\partial}{\partial\gamma} }\pi(\theta) \\
&=-\int_0^T{ \scriptstyle \frac{\partial}{\partial\gamma} }\lambda^*(t;\theta)\dd t + \sum_{0\leq t_i\leq T} \frac{{ \scriptstyle \frac{\partial}{\partial\gamma} }\lambda^*(t_i)}{\lambda^*(t_i)} -{ \scriptstyle \frac{\partial}{\partial\gamma} }\pi(\theta)
\end{aligned}
\]

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_30_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_30_1.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_30_2.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_32_0.png}
    \end{center}
    { \hspace*{\fill} \\}

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_34_0.png}
    \end{center}
    { \hspace*{\fill} \\}
